Still Missing a Cool Title

### 2006-05-30 talk/dls2006/paper.tex
Manual merge of r27925 to r27930 of draft.txt.  At this point the two documents are in sync and only the .tex should be updated.

Applying the expressiveness or at least syntax of very high-level and
dynamically typed languages to their implementation has been
investigated many times.

One typical approach is writing a static compiler.  The viability of
and effort required for such an approach depend usually on the binding
and dispatch semantics of the language.  Common Lisp native compilers,
usable interactively and taking functions or files as compilation
units are a well-known example of that approach.  Late binding for all
names and load semantics make such an approach very hard for Python,
if speed improvements are desired.

It is more relevant to consider and compare with projects using
dynamic and very-high level languages for interpreters and VM
implementations, and Just-In-Time compilers.

[Scheme48] was a Scheme implementation using a restricted Scheme, with
static type inference based on Hindley-Milner, this is viable for
Scheme as base language. Portability and simplicity were its major
goals.

[Squeak] is a Smalltalk implementation in Smalltalk. It uses SLang, a
very restricted subset of Smalltalk with few types and strict
conventions such that mostly direct translation to C is possible.
Both the VM and the object memory and garbage collector support are
explicitly written together in this style. Again simplicity and
portability were the major goals, not sophisticated manipulation and
analysis and waving in of features.

[Jikes RVM] is a Java VM and Just-In-Time compiler written in Java.
Bootstrapping happens by self-applying the compiler on a host VM, and
dumping a snapshot from memory of the resulting native code.

This approach enables directly high performance, at the price of
portability as usual with pure native code emitting
approaches. Modularity of features, when possible, is achieved with
normal software modularity. The indirection costs are taken care by
the inlining done by the compiler, sometimes even through explicit
ways to request for it. In particular this modular approach is used
for implementing a range of choices for GC support XXX ref.  This was
the inspiration for PyPy own GC framework, although much more tuning
and work went into Jikes RVM. PyPy own GC framework also exploits
inlining of helpers and barriers to recover performance.

Jikes RVM native JIT compilers can likely not easily be retargeted to
run on top and target another VM (for example a CLR runtime) instead
of hardware processors. Also Jikes RVM pays the complexity of writing
a JIT up-front, which also means that features and semantics of the
language are encoded in the JIT compiler code, meaning likely that
major changes would correspond to major surgery needed on it.

PyPy more indirect approach, together hopefully with our future work
on generating a JIT compiler, tries to overcome these limitations, at
the price of some more effort required to achieve very good
performance. It is too soon to compare completely the complexity (and
performance) trade-offs of these approaches.

XXX Jython, IronPython, UVM.

@@

XXX

nice interpreter not polluted by implementation details
here Python, but any interpreter works

architecture allows implementing features at the right level

dynamic language enables defining our own various type systems
[ref pluggable type systems]

practical VMs will result with a bit more efforts

>>>>

Applying the expressiveness - or at least the syntax - of very
high-level and dynamically typed languages to their implementation has
been investigated many times.

One typical approach is writing a static compiler.  The viability of,
and effort required for, such an approach depends usually on the binding
and dispatch semantics of the language.  Common Lisp native compilers,
usable interactively and taking single functions or files as compilation
units, are a well-known example of that approach.  However, the late
binding for all names and the load semantics make such an approach very
hard for Python, if speed improvements are desired.

In this context, it is more relevant to consider and compare with
projects using dynamic and very-high-level languages for interpreters
and VM implementations, and Just-In-Time compilers.

Scheme48 was a Scheme implementation using a restricted Scheme, PreScheme
[PreScheme], with static type inference based on Hindley-Milner.  This is
viable for Scheme as base language.  Simplicity and portability across C
platforms were its major goals.

Squeak [Squeak] is a Smalltalk implementation in Smalltalk. It uses SLang, a
very restricted subset of Smalltalk with few types and strict
conventions, which can be mostly directly translated to C.
The VM, the object memory and the garbage collector support are
explicitly written together in this style. Again simplicity and
portability were the major goals, as opposed to sophisticated manipulation
and analysis or "weaving" in of features as transformation aspects.

Jikes RVM [DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00] is a
Java VM and Just-In-Time compiler written in Java.
Bootstrapping happens by self-applying the compiler on a host VM, and
dumping a snapshot from memory of the resulting native code.
This approach directly enables high performance, at the price of
portability - as usual with pure native code emitting
approaches. Modularity of features, when possible, is achieved with
normal software modularity. The indirection costs are taken care of by
the compiler performing inlining (which is sometimes even explicitly
requested).  In particular this modular approach is used
for implementing a range of choices for GC support [Jikes-GC].  This was
the inspiration for PyPy's own GC framework, although much more tuning
and work went into Jikes RVM.  PyPy's own GC framework also exploits
inlining of helpers and barriers to recover performance.

Jikes RVM's native JIT compilers [Jikes-JIT]
are not meant to be retargetted to run in other environments
that hardware processors, for example in a CLR/.NET
runtime. Also Jikes RVM pays the complexity of writing
a JIT up-front, which also means that features and semantics of the
language are encoded in the JIT compiler code.  Major changes of the
language are likely to correspond to major surgery of the JIT.

PyPy's more indirect approach, together hopefully with our future work
on generating a JIT compiler, tries to overcome these limitations, at
the price of some more efforts required to achieve very good
performance. It is too soon for a complete comparison of the complexity,
performance and trade-offs of these approaches.

@@

The PyPy project aims at showing that dynamic languages are suitable and
quite useful for writing virtual machines in.  We believe that we have
achieved this objective.  The present paper gave an overview of the
architecture that enabled this result.  Experiments suggest that
practical virtual machines could reasonably follow in the near future,
with faster-than-current virtual machines with JIT specialization
techniques for the mid-term future.

Targetting platforms that are very different from C/Posix is work in
progress, but given that many of the initial components are shared with
the existing stack of transformations leading to C, we are confident
that this work will soon give results.  Moreover, we believe that these
results will show a reasonable efficiency, because the back-ends for VMs
like Squeak and .NET can take advantage of high-level input (as opposed
to trying to translate, say, C-like code to Smalltalk).

A desirable property of our approach is to allow a given language and VM
to be specified only once, in the form of an interpreter.  Moreover, the
interpreter can be kept simple (and thus keep its role as a
specification): not only is it written in a high-level language, but it
is not overloaded with low-level design choices and implementation
details.  This makes language evolution and experimentation easier.
More generally, this property is important because many interpreters for
very difference languages can be written: the simpler these interpreters
can be kept, the more we win from our investment in writing the
tool-chain itself - a one-time effort.

Dynamic languages enable the definition of multiple custom type systems,
similar to \textit{pluggable type systems} in [Bracha] but with simple
type inference instead of explicit annotations.  This proved a key
feature in implementing our translation tool-chain, because it makes a
many-levels approach convenient: each abstraction level can provide an
implementation for some of the features that the higher levels
considered primitive.  It offsets the need to define a minimal kernel of
primitives and build everything on top of it; instead, we have been able
to implement, so to speak, the right feature at the right level.
